
<!-- --- -->
<!-- bibliography: refs.bib -->
<!-- csl: apa.csl -->
<!-- --- -->

<!-- {{< include _extensions/_maths_shortcuts.qmd >}} -->

```{r}
#| include: false
library(tidyverse)
library(kableExtra)
library(gt)
```


### Ordinal data

Consider the case of analysing multivariate data $\mathbf y = (y_{1}, \ldots, y_{p})^\top$, where each item $y_{i}$ is an ordinal random variable with $m_i$ categories, $i=1,\dots,p$.
Let $\mathcal R = \{ \mathbf c = (c_1,\dots, c_p)^\top \mid c_i \in \{1,\dots, m_i\}\}$ be the set of all possible response patterns, and let $R=\prod_{i} m_i$ be the cardinality of this set.
The joint probability of observing a response pattern $\mathbf c_r \in \mathcal R$ is given by
$$
\pi_r = \Pr(\mathbf y = \mathbf c_r) = \Pr(y_1 = \mathbf c_{r1}, \ldots, y_p = \mathbf c_{rp}), \hspace{2em} r = 1, \ldots, R,
$$
with $\sum_r \pi_R = 1$.
Collect all response probabilities into the vector $\boldsymbol \pi = (\pi_1, \ldots, \pi_R)^\top \in [0,1]^R$.
An example with $p=3$, $m_1=2$, and $m_2=m_3=3$ is given below.
In total, there are $R=2 \times 3 \times 3 = 18$ response patterns as shown in @tbl-response-patterns.

```{r}
#| include: false

tab_rp <-
  expand_grid(
    y1 = 1:2,
    y2 = 1:3,
    y3 = 1:3
  ) |> 
  unite("pattern", everything(), sep = "", remove = FALSE) |>
  mutate(r = row_number()) |>
  select(r, starts_with("y"), pattern)
```

::: {#tbl-response-patterns layout-ncol=2}

```{r}
#| html-table-processing: none
#| echo: false
tab_rp |>
  slice(1:9) |>
  gt() |>
  cols_label(
    r = md("$r$"),
    y1 = md("$y_1$"),
    y2 = md("$y_2$"),
    y3 = md("$y_3$"),
    pattern = "Pattern"
  ) |>
  tab_options(table.width = "80%")
```

```{r}
#| html-table-processing: none
#| echo: false
tab_rp |>
  slice(-(1:9)) |>
  gt() |>
  cols_label(
    r = md("$r$"),
    y1 = md("$y_1$"),
    y2 = md("$y_2$"),
    y3 = md("$y_3$"),
    pattern = "Pattern"
  ) |>
  tab_options(table.width = "80%")
```

Response patterns for $p=3$ with $m_1=2$, and $m_2=m_3=3$.
:::

Later on we wish to use lower-order residuals to assess the fit of a model to the data, which first requires a description of lower-order moments and its connection to the joint response probabilities.
<!-- Let $I_{ik} = [y_i = k]$ be the indicator variable for the event that $y_i$ takes the value $k$, where $[\cdot]$ is the Iverson bracket. -->
Marginally, each $y_i$ can be viewed as a multinoulli random variable with event probabilities $\pi^{(i)}_k = \Pr(y_i = k)$, $k=1,\dots m_i$, that sum to one.
Therefore, this univariate distribution is characterised by its $(m_i-1)$ *moments* $\pi^{(i)}_2,\dots,\pi^{(i)}_{m_i}$, with the first moment being redundant due to the sum to unity constraint.
All univariate moments can be collected into the vector $\dot{\boldsymbol\pi}_1 = (\pi^{(i)}_k)^\top$ whose dimension is $S_1 = \sum_i (m_i-1)$.
In a similar light, the bivariate distribution of $(y_i, y_j)$ is characterised by its $(m_i-1)(m_j-1)$ *joint moments* $\pi^{(ij)}_{k,l} = \Pr(y_i = k, y_j = l)$, $k=2,\dots,m_i$, $l=2,\dots,m_j$.
Also collect all bivariate moments into the vector $\dot{\boldsymbol\pi}_2 = (\pi^{(ij)}_{k,l})^\top$ whose dimension is $S_2 = \sum_{i<j} (m_i-1)(m_j-1)$.
Finally, denote by $\boldsymbol\pi_2 = (\dot\bpi_1^\top, \dot\bpi_2^\top)^\top$ the vector of multivariate moments up to order 2, which is a vector of length $S = S_1 + S_2$.

Because the lower order moments are contained in the higher order moments, the vector $\boldsymbol\pi_2$ can be extracted from the joint probabilities $\bpi$ via a linear operation $\bpi_2 = \bT_2 \bpi$ [@jamil2025pairwise].
As an example, continuing from the $p=3$ instance above, the moments for the first variable $y_1$, $\Pr(y_1=2)$ can be obtained by *summing* over all joint probabilities whose patterns contain $y_1=2$.
The positions of these joint probabilities in the vector $\bpi$ are picked up by the first row of the matrix $\bT_2$.
Similarly, the two bivariate moments of $(y_1,y_2)$, i.e. $\pi^{(12)}_{22}$ and $\pi^{(12)}_{23}$ are obtained by summing over the joint probabilities whose patterns contain $y_1=2$ and $y_2=2$, and $y_1=2$ and $y_2=3$, respectively.

::: {#fig-T2-matrix}
```{r}
#| echo: false
options(width = 100)
create_T2_mat <- function(m) {
  # m: integer vector of length p, where m[i] = number of categories of variable i
  p <- length(m)
  # 1) all joint patterns (rows = âˆ m[i], cols = p)
  patterns <- expand.grid(rev(lapply(m, seq_len)), KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE)
  patterns <- patterns[, rev(seq_len(p))] # reverse to match y1, y2, ...
  n_pat <- nrow(patterns)
  
  # 2) precompute total number of rows: sum_i (m[i]-1) + sum_{i<j} (m[i]-1)*(m[j]-1)
  uni_rows <- sum(m - 1)
  biv_rows <- 0L
  for(i in seq_len(p-1)) for(j in (i+1):p)
    biv_rows <- biv_rows + (m[i]-1)*(m[j]-1)
  total_rows <- uni_rows + biv_rows
  
  # 3) allocate
  out <- matrix(0L, nrow = total_rows, ncol = n_pat)
  rn  <- character(total_rows)
  
  # 4) fill univariate indicator rows
  r <- 1L
  for(i in seq_len(p)) {
    for(cat in 2:m[i]) {
      out[r, ] <- as.integer(patterns[[i]] == cat)
      rn[r]   <- paste0("Y", i, "=", cat)
      r       <- r + 1L
    }
  }
  
  # 5) fill bivariate indicator rows
  for(i in seq_len(p-1)) for(j in (i+1):p) {
    for(c1 in 2:m[i]) for(c2 in 2:m[j]) {
      out[r, ] <- as.integer(patterns[[i]] == c1 & patterns[[j]] == c2)
      rn[r]   <- paste0("Y", i, "=", c1, ",Y", j, "=", c2)
      r       <- r + 1L
    }
  }
  
  rownames(out) <- rn
  colnames(out) <- apply(patterns, 1, paste0, collapse = "")
  out
}
create_T2_mat(c(2, 3, 3))
```

Matrix $\bT_2$ for the case of $p=3$ with $m_1=2$, and $m_2=m_3=3$.

:::

Note that this construction of lower-order moments generalises to any order $q \le p$, but the total number of moments up to order $q$ grows combinatorially in both $p$ and the category counts $m_i$, yielding design matrices $\mathbf{T}_q$ that can become computationally burdensome. 
Moreover, although we arbitrarily dropped the first moment in the foregoing construction, the choice of which category to omit is immaterial.
This is because category probabilities sum to one, so excluding any one category produces a similar-dimensional parameterisation algebraically equivalent to excluding any other. 
For further details, consult @reiser1996analysis and @maydeu2006limited.

### Confirmatory factor analysis

### Parameter estimation

### Distribution of residuals

### Wald type tests

### Pearson type tests

### General GOF tests

### Estimation of degrees of freedom