---
title: Pairwise Maximum Likelihood
author: Haziq Jamil
grid:
  sidebar-width: 200px
  body-width: 900px
  margin-width: 0px
  gutter-width: 1.5em
editor_options: 
  chunk_output_type: console
execute:
  cache: true
---


```{r}
#| include: false
here::i_am("notebooks/pml.qmd")
source(here::here("notebooks/ligof.R"))
```

We can use the `estimator = "PML"` argument in `{lavaan}` to fit a model using pairwise maximum likelihood (PML) estimation. 

```{r}
fit <- cfa("f1 =~ y1 + y2 + y3 + y4 + y5", data = dat, estimator = "PML", 
           std.lv = TRUE)
coef(fit)
```

Alternatively, we can use the addon package `{lavaan.pl}` for more efficient PML estimation.

```{r}
library(lavaan.pl)
# fit <- lavaan.pl::cfa("f1 =~ y1 + y2 + y3 + y4 + y5", data = dat, std.lv = TRUE)
coef(fit)
```

## $\mathbf T_2$ matrix

This is the design matrix that maps all joint probabilities to univariate and bivariate moments.

```{r}
CREATE_T2_MAT <- function(m) {
  # m: integer vector of length p, where m[i] = number of categories of variable i
  p <- length(m)
  # 1) all joint patterns (rows = âˆ m[i], cols = p)
  patterns <- expand.grid(rev(lapply(m, seq_len)), KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE)
  patterns <- patterns[, rev(seq_len(p))] # reverse to match y1, y2, ...
  n_pat <- nrow(patterns)
  
  # 2) precompute total number of rows: sum_i (m[i]-1) + sum_{i<j} (m[i]-1)*(m[j]-1)
  uni_rows <- sum(m - 1)
  biv_rows <- 0L
  for(i in seq_len(p-1)) for(j in (i+1):p)
    biv_rows <- biv_rows + (m[i]-1)*(m[j]-1)
  total_rows <- uni_rows + biv_rows
  
  # 3) allocate
  out <- matrix(0L, nrow = total_rows, ncol = n_pat)
  rn  <- character(total_rows)
  
  # 4) fill univariate indicator rows
  r <- 1L
  for(i in seq_len(p)) {
    for(cat in 2:m[i]) {
      out[r, ] <- as.integer(patterns[[i]] == cat)
      rn[r]   <- paste0("Y", i, "=", cat)
      r       <- r + 1L
    }
  }
  
  # 5) fill bivariate indicator rows
  for(i in seq_len(p-1)) for(j in (i+1):p) {
    for(c1 in 2:m[i]) for(c2 in 2:m[j]) {
      out[r, ] <- as.integer(patterns[[i]] == c1 & patterns[[j]] == c2)
      rn[r]   <- paste0("Y", i, "=", c1, ",Y", j, "=", c2)
      r       <- r + 1L
    }
  }
  
  rownames(out) <- rn
  colnames(out) <- apply(patterns, 1, paste0, collapse = "")
  out
}
CREATE_T2_MAT(rep(2, 5))
```


## Influence matrix

To estimate the influence matrix, we need a couple of ingredients:

- Inverse Hessian matrix evaluated at the PML estimates
- Jacobian $\partial \boldsymbol\pi_{\text{pair}}(\theta) / \partial\theta$ evaluated at the PML estimates
- Weight matrix $\operatorname{diag}(\boldsymbol\pi_{\text{pair}}(\theta))^{-1}$
- $\boldsymbol G$ transformation matrix which brings $\boldsymbol\pi_{\text{pair}} \mapsto \boldsymbol \pi$

```{r}
# Inverse Hessian
Hinv <- lavaan:::lav_model_information_observed(
  lavmodel = fit@Model,
  lavsamplestats = fit@SampleStats,
  lavdata = fit@Data,
  lavoptions = fit@Options,  
  lavcache = fit@Cache, 
  lavimplied = fit@Implied,
  inverted = TRUE
)
round(Hinv, 3)

# Jacobian -- for PML, there are 4 x (5C2) = 40 rows corresponding to pairwise
# probabilities.
Delta <- lavaan.bingof:::get_Delta_mats(fit)$Delta_til
round(Delta, 3)

# Weight matrix
W <- diag(1 / unlist(lavaan:::lav_tables_pairwise_model_pi(fit)))
diag(W)

# Transformation matrix -- in Jamil et al. (2025), this is the design matrix
# called B such that G = B %*% T2
B  <- lavaan.bingof:::Beta_mat_design(nvar = 5)
T2 <- lavaan.bingof:::create_T2_mat(p = 5)
# G <- B %*% T2
# Matrix::Matrix(G, sparse = TRUE)
G2 <- B
Matrix::Matrix(G2, sparse = TRUE)

# Influence matrix
Q <- Hinv %*% t(Delta) %*% W %*% G2
round(Q, 3)

tinytest::expect_equal(
  Q,
  lavaan.bingof:::test_begin(fit, NULL, FALSE)$B2
)
```

## $\boldsymbol\Omega_2$ matrix

This is the (asymptotic) variance-covariance matrix of the univariate and bivariate moments.

```{r}
Delta2 <- create_Delta2_matrix(fit)
Sigma2 <- lavaan.bingof:::create_Sigma2_matrix(fit)
D2Q <- Delta2 %*% Q

Omega2 <- 
  Sigma2 -
  D2Q %*% Sigma2 -
  Sigma2 %*% t(D2Q) +
  D2Q %*% Sigma2 %*% t(D2Q)

Matrix::Matrix(Omega2) |> 
  round(3)

tinytest::expect_equal(
  Omega2,
  lavaan.bingof:::test_begin(fit, NULL, FALSE)$Omega2
)
```

## Wald and Pearson test


```{r}
ligof_test <- function(lavobject) {
  
  n <- nobs(lavobject)
  Delta2 <- create_Delta2_matrix(lavobject)
  # Delta2 <- lavaan.bingof:::get_Delta_mats(lavobject)$Delta2
  Sigma2 <- lavaan.bingof:::create_Sigma2_matrix(lavobject, method = "theoretical")
  uni_bi_moments <- lavaan.bingof::get_uni_bi_moments(lavobject)
  p2_hat <- with(uni_bi_moments, c(pdot1, pdot2))
  pi2_hat <- with(uni_bi_moments, c(pidot1, pidot2))
  e2_hat <- p2_hat - pi2_hat

  # Build influence matrix
  Hinv <- lavaan:::lav_model_information_observed(
    lavmodel = lavobject@Model,
    lavsamplestats = lavobject@SampleStats,
    lavdata = lavobject@Data,
    lavoptions = lavobject@Options,  
    lavcache = lavobject@Cache, 
    lavimplied = lavobject@Implied,
    inverted = TRUE
  )
  Delta <- lavaan.bingof:::get_Delta_mats(lavobject)$Delta_til
  W <- diag(1 / unlist(lavaan:::lav_tables_pairwise_model_pi(lavobject)))
  G2 <- lavaan.bingof:::Beta_mat_design(nvar = 5)
  Q <- Hinv %*% t(Delta) %*% W %*% G2
  
  # Build Omega2 matrix
  D2Q <- Delta2 %*% Q
  Omega2 <- 
    Sigma2 -
    D2Q %*% Sigma2 -
    Sigma2 %*% t(D2Q) +
    D2Q %*% Sigma2 %*% t(D2Q)
  Omega2 <- (Omega2 + t(Omega2)) / 2

  out <- list()
  
  # Wald test
  Xi <- MASS::ginv(Omega2)
  X2 <- n * colSums(e2_hat * (Xi %*% e2_hat))
  df <- Matrix::rankMatrix(Omega2) - ncol(Delta2)
  pval <- pchisq(X2, df, lower.tail = FALSE)
  out$Wald <- tibble(X2 = X2, df = df, pval = pval)
  # out$WaldLB <- lavaan.bingof::Wald_test(lavobject) |> select(X2, df, pval)
  
  # Pearson test
  Xi <- diag(1 / pi2_hat)
  X2 <- n * colSums(e2_hat * (Xi %*% e2_hat))
  df <- nrow(Delta2) - ncol(Delta2)
  out$Pearson <- 
    lavaan.bingof:::moment_match(X2, Xi, Omega2, df, 3) |>
    as_tibble()
  out$Pearson$pval <- pchisq(out$Pearson$X2, df, lower.tail = FALSE)
  # out$PearsonLB <- lavaan.bingof::Pearson_test(lavobject) |> select(X2, df, pval)

  out
}

ligof_test(fit)
```

## Verification

```{r}
library(furrr)
plan(multisession)
B <- 2000  # no. of simulations

sim_fun <- function(i) {
  set.seed(i)
  dat <- simulate_data(n = 1000)
  fit <- lavaan::cfa("f1 =~ y1 + y2 + y3 + y4 + y5", dat, std.lv = TRUE, estimator = "PML")
  ligof_test(fit) |>
    bind_rows(.id = "name") |>
    mutate(sim = i, name = factor(name, levels = c("Wald", "Pearson"))) |>
    select(sim, everything())
}

res <- future_map(seq_len(B), sim_fun, .progress = TRUE,
                  .options = furrr_options(seed = TRUE))
```

```{r}
bind_rows(res) |>
  mutate(
    df = mean(df),
    observed = sort(X2),
    expected = qchisq(ppoints(n()), df = df),
    .by = name
  ) |>
  ggplot(aes(expected, observed, col = name)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "QQ-plot of M2 against chi-square distribution with 5 df",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 20)) +
  theme_bw()
```

```{r}
bind_rows(res) |>
  ggplot(aes(x = X2, y = ..density.., fill = name)) +
  geom_histogram(col = "gray90") +
  geom_line(
    data = 
      bind_rows(res) |>
      group_by(name) |>
      summarise(df = mean(df), .groups = "keep") |>
      mutate(
        x = list(seq(0, 20, 0.1)),
        y = map(x, \(x) dchisq(x, df = first(df))),
        .groups = "drop"
      ) |>
      unnest(c(x, y)),
    aes(x, y),
    linewidth = 1
  ) +
  facet_wrap(name ~ ., scales = "free_y") + 
  labs(fill = NULL) +
  theme_bw()
```

```{r}
bind_rows(res) |>
  ggplot(aes(x = pval, y = ..density.., fill = name)) +
  geom_histogram(binwidth = 0.05, boundary = 0, col = "white") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(name ~ ., scales = "free_y") + 
  labs(fill = NULL) +
  theme_bw()
```




